Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-22 text/manuscript-full-KS.qmd
went through discussion after karina

---
title: "Structuring College Access: The Market Segment Model and College Board Geomarkets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)", "Crystal Han (Stanford University)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

The discussion of homophily by @RN4975 describes not only the logic of The Market Segment model but also a broader institutional logic that characterizes the enrollment management industry. Homophily argues that there is a natural cultural match between students and colleges, suggesting colleges should find prospective students who resemble students who enrolled at their institution in the past and focus recruiting efforts on the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The study also finds great overlap in the schools visited by Villanova University and the University of Notre Dame, with a disproportionate focus on visits to Catholic schools. <!--# When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. --> As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

The Market Segment Model begins with the status attainment model, that student demand for college is a function of parental education and parental income. @RN4982 [chapter 3] illustrated this argument by demonstrating a correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that that households of specific class characteristics and college aspirations are likely to live in particular communities within metropolitan areas and that these geographic territories can be meaningfully captured by Geomarkets. Although Geomarkets appear coarse by contemporary standards, they were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment (i.e., local, in-state, regional, or national), identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), a standard ouput in EPS software. When considering new Geomarkets to target, EPS recommends colleges utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A).

@@

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. <!--#  [KARINA SUMMARIZE FINDINGS IN A COUPLE SENTENCES; REVISE PARAGRAPH AS NECESSARY]--> Findings illustrate that in 1980, when Geomarkets were being created, Black and Hispanic residents tended to be highly concentrated in a single, usually the poorest Geomarket in the Chicago, Dallas-Fortworth, and to some extent the Los Angeles metropolitan areas. These Geomarket patterns extend to other metropolitan areas in the country, such as CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), and PA5 -- Philadelphia County. By 2020, various Geomarket landscapes emerged. In Chicago, Black residents remain concentrated in the poorest Geomarket. In contrast, in Dallas-Fortworth, Black and Hispanic communities are distributed across multiple Geomarkets; however, significant socioeconomic disparities persist between Geomarkets. The Los Angeles metropolitan area presents another distinctive Geomarket landscape in which Black, Hispanic, and Asian communities across a range of socioeconomic levels are distributed across multiple Geomarkets rather than concentrated in one or two. <!--# [KARINA -- ONE OR TWO SENTENCES] -->

These findings are not surprising given the extent of race and class segregation in the US. Furthermore, the correlation between Geomarkets and class is expected because Geomarket borders were created in service of conceptual model that views demand for higher education as a function of class. @RN4982 is written around New England, with a particular focus on Geomarkets in Connecticut and Massachusetts. @RN4982 do not discuss the intersection of student demand and race. Although unsurprising for early 1980s social science, this silence is conspicuous considering the salience of political controversy surrounding "busing" in Boston and nearby metropolitan areas.<!--# KS: COULD CUT THIS IF NEEDED; MAIN TAKEAWAY IS JUST THAT RQ1 IS UNSURPRINSING EMPIRICALLY VIA SEGREGATION AND THEORETICALLY VIA ZEMSKY & OEDEL-->

Research question 2 analyzes the racial and socioeconomic composition of student list purchases. We examine purchases that include all Geomarkets in a metropolitan area in order to assess the potential consequences if including or excluding particular Geomarkets from the purchase. This question is motivated by the @RN4982 recommendation that selective colleges that enroll students from the regional and national market segment should target affluent Geomarkets, while community colleges and non-selective 4-year colleges should focus on middle- and working-class Geomarkets. Findings across metropolitan cases suggest student list orders that use Geomarket filters to exclude middle- and working-class Geomarkets - such as IL11 -- City of Chicago, TX19 -- City of Dallas, TX20 -- City of Fort Worth, CA19 -- East Los Angeles, and CA21 -- South and South Central Los Angeles, in targeting high-achieving students may still capture some Black, Hispanic, and Asian prospects. However, these prospects are far more unlikely to be first-generation students of color. <!--# KARINA -- ADD TEXT SUMMARIZING RESULTS.-->

This article adds a new perspective to existing explanations about how students are sorted into colleges. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. The Market Segment Model is the status attainment model applied to geodemographic market research \[CITE\]. @RN5033 finds that the educational achievement of Asian American students whose parents are immigrants tend to have high educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Drawing from @RN5033, one explanation for this finding is that Asian students living in low-income Geomarkets tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents who were born in the U.S. <!--# KARINA NOTES -- USE RESULTS TO SPEAK TO MAJOR THEORIES.-->

The cultural capital model explains a process by which upper and upper-middle class families sort themselves into selective colleges by providing their children with the pedigree (academic opportunities, extracurriculars) valued by selective colleges, information about how to navigate the admissions gauntlet, and social networks that provide an inside track [@RN1228; @RN1241; @RN4991; @RN1814]. This model explains how affluent households maintain a disproportionate enrollment share at selective institutions in an era of holistic admissions ostensibly designed to increase racial and class diversity [@RN4991]. Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of mobility while maintaining a much larger flow of intergenerational class transmission [@RN861]. @RN4722 -- who obtained parental income from federal income tax returns for every U.S. college -- show that the disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. These patterns are consistent with the observation by @RN1297 [p. 241-242] that educational credentials are property that "support their holders'\ldots claims to monopolize socially and economically advantageous positions."

Supply-side explanations of credentialism and more recent scholarship on enrollment management complement the cultural capital model of sorting students to colleges. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to determine the competition for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869] and this credentials arms race favors affluent households. Scholarship on enrollment management describes which students colleges want to enroll and what colleges do to attract these students at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. Counselors at well-resourced schools, especially private schools, are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognzing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The orchestra needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023; @RN4758]. However, these analyses conceive of recruiting as something that is done solely by individual colleges.

Existing explanations of how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. @RN4982 created the Market Segment Model and Geomarkets based on a snapshot of student demand from 1980. @RN4982 argues that demand for higher education is correlated with class, ignoring the historical structural barriers that produced class- and race-based inequality in 1980 student demand. In itself, the Market Segment Model is an unremarkable social science depiction. However, the College Board inscribed Geomarkets and the Market Segment Model into EPS software that tells colleges which Geomarkets and high schools to target. Technologies that target customers based on geography inevitably leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. Furthermore, technologies that use the logic of predictive analytics -- identify correlates in past cases to make predictions for future cases -- are prone to "pernicious feedback loops" [@RN4849]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused historic place-based inequality in student demand are amplified because EPS recommends that colleges focus recruiting resources on localities that already have high student demand for peer colleges. Geomarkets were subsequently incorporated into the Student Search Service product. We show that that excluding low-income, non-white Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-white students with strong test scores. <!--# KS Notes: THIS IS THE STRONGEST PARAGRAPH -->

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field salient to college access. @RN527 \[p. 148\] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities. College access scholarship from economics often analyzes the effects of district, state, and federal [e.g., @RN5043] policies and the effects of information interventions [@RN4018]. These lines of research do not consider the growing role of third-party vendors, which are increasingly owned by private equity interests and offer software-as-service platforms that perform core functions for schools and colleges.^[@RN4991 describes the role of independent admissions consultants hired by families as third-party vendors that engage on the demand-side.]

@list_biz describes four key dynamics in the enrollment management (EM) industry, with a focus on the market for student list data. First, EM consulting firms are central to the creation and implementation of recruiting campaigns [@RN5027]. The top two EM consultancies are Ruffalo Noel Levitz, which claims to serve 1,900 colleges annually [@rnlabout_2025], and EAB, which claims to serve 2,100 colleges [@about_eab]. College reliance on EM consultancies is partially explained by the high level of burnout and employee turnover in the EM profession [@hoover_em]. When we issued records requests to public universities about their student list purchases, at least 50% of universities indicated that they outsourced student list purchases to an EM consultancy. In many cases, the university was buying names each year but no university employee had knowledge about which vendors they were buying from or which search filters they were utilizing.

Second, technological advances creates new means of identifying and serving prospects. College Board and ACT leveraged data on test-takers to enjoy a near duopoly in market for student list data, one that lasted for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded new sources of student list data. Another data source is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than 10 million K-12 students and by 40% of US high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based targeting.

The third dynamic in the EM industry has been the growth of test-optional admissions movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company \[CITE\], failing \[\], and being acquired by Nexus Capital in 2024 \[CITE\]. Over the last two decades, the testing companies developed new search filters that explicitly incorporated predictive analytics -- for example College Board's Geodemographic Segment filters \[CITE\] and ACT's predicted probability of enrollment filter \[CITE\] -- developed free college search engines, and entered the enrollment management consulting market more aggessively. College Board has weathered the test-optional movement thanks to robust revenues from the Advanced Placement product, but the decline in PSAT/SAT/PreACT/ACT test-takers has undermined their oligopoly in the market for student list data.

The fourth dynamic is the transformation from owner-operated to private-equity owned firms and the erosion of the distinction between consultancy and software vendor. The 1980s and 1990s was a period of market entry in EM, when college admissions professionals or professors decided to hang up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. After regulatory scrutiny pushed private-equity interests out of the for-profit college market [@RN5022], private equity investors found value in acquiring firms that provided services to direct providers. Drawing from resource dependence theory, @list_biz describe the proliferation of horizontal acquisitions to grow market share and vertical acquisitions designed to make customers more dependent on particular recruiting products.

EAB exemplifies the the trend towards private equity ownership and industry concentration. EAB is the preeminent EM consultancy, offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB trace to Bill Royall, who founded Royall & Company in 1983 to provide direct marketing and fundraising for Republican political campaigns [@jump_2020]. Royall did not sign its first colege client for several years, but EM became the primary focus by 1995. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

As an owner-operated consultancy, Royall was the market leader in purchasing student lists from student list vendors on behalf of clients. Under Vista, EAB entered the market for student list data by completing a series of acquisitions that targeted the "leads" and "inquiries" stages of the enrollment funnel. In 2019, EAB acquired YouVisit, a leading provider of virtual tours. YouVisit is an engine for "inquiries" in that taking a virtual tour shows interest in a college and leaves a trail of contact information for subsequent recruiting interventions. In 2020, EAB acquired Cappex, a leading free college search engine, which is an engine for leads in that students provide contact details and other information. In 2021, EAB and PowerSchool -- the leading provider of K12 student information systems and a subsidiary of Vista -- jointly acquired the EM consultancy/software vendor Hobsons. EAB acquired Hobson's student success software, Starfish, while PowerSchool acquired Hobson's Naviance product and its sister recruiting product Intersect. EAB paid PowerSchool to become the exlusive reseller of Intersect. Finally, in 2022, EAB acquired Concourse, the leading "direct admissions" platform. EAB integrated these acquisitions to create the [Enroll360](https://eab.com/solutions/enroll360/) recruiting platform.^[The Enroll360 press release states, "We spent the last couple of years creating a connected recruitment ecosystem that allows enrollment leaders to keep pace with students pursuing increasingly digital journeys to college. This work led us to join forces with several leading companies: Cappex, Intersect, Wisr, and YouVisit\ldots Individually, each solution can solve important challenges at various stages of the enrollment funnel\ldots By bringing these capabilities together, our vision is to reinvent how enrollment leaders reach their goals [@koppenheffer_2021]."]

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50 for each "name"). The goal was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software, which is more expensive than purchasing names. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board's *Student Search Service,* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and wheather students used Naviance to "research competitor institutions" [@markup_naviance]. <!--#  A procurement document from the University of Utah explains that the Intersect subscription is necessary to recruit the proprietary database of Naviance users: [@utah_procurement]:-->

> <!--# The marketing services being purchased operates within the PowerSchool Naviance platform, which is a proprietary system, that no other company has access to. There is unique group of prospective students who are only in the PowerSchool Naviance platform. We have spoken with several vendors where we can reach out to prospective students. Each vendor has their own proprietary website or database where select students can be reached.-->

>>>>

---
title: "Structuring College Access: The Market Segment Model and College Board Geomarkets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)", "Crystal Han (Stanford University)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  #docx:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

The discussion of homophily by @RN4975 describes not only the logic of The Market Segment model but also a broader institutional logic of the enrollment management industry. The logic of homophily assumes that there is a natural cultural match between students and colleges, suggesting colleges should find prospective students who resemble students who enrolled at their institution in the past and focus recruiting efforts on the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. <!--# The study also finds great overlap in the schools visited by Villanova University and the University of Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. -->  As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

The Market Segment Model begins with the status attainment model idea that student demand for college is a function of parental education and parental income. @RN4982 illustrated this argument by demonstrating a correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that that households of specific class characteristics and college aspirations are likely to live in particular communities and that these geographic territories can be meaningfully captured by Geomarkets. Although Geomarkets appear coarse by contemporary standards, they were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment (i.e., local, in-state, regional, or national), identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), a standard ouput in EPS software. When considering new Geomarkets to target, EPS recommends colleges utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A).

@@

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. Findings illustrate that in 1980, when Geomarkets were being created, Black and Hispanic residents tended to be highly concentrated in the poorest Geomarket in the Chicago, Dallas-Fortworth, and to some extent the Los Angeles metropolitan areas. These Geomarket patterns extend to other metropolitan areas in the country, such as CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), and PA5 -- Philadelphia County. In 2020 Chicago, Black residents remain concentrated in the poorest Geomarket but Gentrification also pushed Black residents to the South and Southwest Suburbs Geomarket. In Dallas-Fortworth, Black and Hispanic communities are distributed across multiple Geomarkets; however, significant socioeconomic disparities persist between Geomarkets. The Los Angeles metropolitan area presents another distinctive Geomarket landscape in which Black, Hispanic, and Asian communities across a range of socioeconomic levels are distributed across multiple Geomarkets rather than concentrated in one or two.

<!--#  These findings are not surprising given the extent of race and class segregation in the US. Furthermore, the correlation between Geomarkets and class is expected because Geomarket borders were created in service of conceptual model that views demand for higher education as a function of class. @RN4982 is written around New England, with a particular focus on Geomarkets in Connecticut and Massachusetts. @RN4982 do not discuss the intersection of student demand and race. Although unsurprising for early 1980s social science, this silence is conspicuous considering the salience of political controversy surrounding "busing" in Boston and nearby metropolitan areas.KS: COULD CUT THIS IF NEEDED; MAIN TAKEAWAY IS JUST THAT RQ1 IS UNSURPRINSING EMPIRICALLY VIA SEGREGATION AND THEORETICALLY VIA ZEMSKY & OEDEL-->

Research question 2 analyzes the racial and socioeconomic composition of student list purchases. We examine purchases that include all Geomarkets in a metropolitan area in order to assess the potential consequences of including or excluding particular Geomarkets from the purchase. This question is motivated by the @RN4982 recommendation that selective colleges that enroll students from the regional and national market segment should target affluent Geomarkets, while community colleges and non-selective 4-year colleges should focus on middle- and working-class Geomarkets. Findings across metropolitan cases demonstrate that if student list purchases exclude working- and middle-class Geomarkets - such as IL11 -- City of Chicago, TX19 -- City of Dallas, TX20 -- City of Fort Worth, CA19 -- East Los Angeles, and CA21 -- South and South Central Los Angeles -- the resulting student list purchases disproportionately exclude Black, and Hispanic students who meet the requisite academic achievement criteria. Although these purchases would still capture some Black, Hispanic, and Asian prospects, these prospects are much less likely to be first-generation students.

This article adds a new perspective to existing explanations about how students are sorted into colleges. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. We conceive of the the Market Segment Model is the status attainment model applied to geodemographic market research, akin to the rise of credit-scoring in UK retail banking [@RN4880]. @RN5033 finds that the educational achievement of Asian American students whose parents are immigrants tend to have high educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Building on @RN5033, Asian students living in low-income Geomarkets may tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents born in the U.S.

The cultural capital model explains a process by which upper and upper-middle class families sort themselves into selective colleges by providing their children with the pedigree (academic opportunities, extracurriculars) valued by selective colleges, information about how to navigate the admissions gauntlet, and social networks that provide an inside track [@RN1228; @RN1241; @RN4991; @RN1814]. This model explains how affluent households maintain a disproportionate enrollment share at selective institutions in an era of holistic admissions ostensibly designed to increase racial and class diversity [@RN4991]. Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of mobility while maintaining a larger flow of intergenerational class transmission [@RN861]. @RN4722 -- who obtained parental income from federal income tax returns for every U.S. college -- show that the disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. These patterns are consistent with the observation by @RN1297 [p. 241-242] that educational credentials are property that "support their holders'\ldots claims to monopolize socially and economically advantageous positions."

Supply-side explanations of credentialism and more recent scholarship on enrollment management complement the cultural capital model of sorting students to colleges. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to determine the competition for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869] and this credentials arms race favors affluent households. Scholarship on enrollment management describes which students colleges want and how colleges go about attracting these students at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. Counselors at well-resourced -- especially private -- schools, are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognzing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The orchestra needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023; @RN4758]. However, these analyses conceive of recruiting as something that is done solely by individual colleges.

Existing explanations of how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. @RN4982 created the Market Segment Model and Geomarkets based on a snapshot of student demand from 1980. @RN4982 argues that demand for higher education is correlated with class, ignoring the historical structural barriers that produced class- and race-based inequality in 1980 student demand. In itself, the Market Segment Model is a social science depiction. However, the College Board inscribed Geomarkets and the Market Segment Model into EPS software that tells colleges which Geomarkets and high schools to target. Technologies that target customers based on geography inevitably leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. Furthermore, technologies that use the logic of predictive analytics -- identify correlates in past cases to make predictions for future cases -- are prone to "pernicious feedback loops" [@RN4849]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused historic place-based inequality in student demand are amplified because EPS recommends that colleges focus recruiting resources on localities that already have high student demand for peer colleges. Geomarkets were subsequently incorporated into the Student Search Service product. We show that that excluding low-income, non-white Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-white students with strong academic achievement.

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field salient to college access. @RN527[p. 148] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities. College access scholarship from economics often analyzes the effects of district, state, and federal policies [e.g., @RN5043]. These literatures ignore the role of third-party vendors, which are often owned by private equity and offer software-as-service platforms that perform core functions for schools and colleges.^[@RN4991 studies independent admissions consultants, third-party vendors on the demand-side.]

@list_biz describes four key dynamics in the enrollment management (EM) industry, with a focus on the market for student list data. First, EM consulting firms are central to the creation and implementation of recruiting campaigns [@RN5027]. The top two EM consultancies are Ruffalo Noel Levitz, which claims to serve 1,900 colleges annually [@rnlabout_2025], and EAB, which claims to serve 2,100 colleges [@about_eab]. College reliance on EM consultancies is partially explained by the high level of burnout and employee turnover in the EM profession [@hoover_em]. When we issued records requests to public universities about their student list purchases, at least 50% of universities indicated that they outsourced student list purchases to an EM consultancy. Often, the university was buying names each year but no employee had knowledge about which vendors they were buying from or which search filters they were utilizing.

Second, technological advances creates new means of identifying and serving prospects. College Board and ACT leveraged data on test-takers to enjoy a near duopoly in market for student list data, one that lasted for several decades. This business can be described as list-based lead generation based on the direct-mail model [@singer1988]. "Free" college search engines (e.g., Cappex, Niche) yielded new sources of student list data. Another data source is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than 10 million K-12 students and by 40% of US high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform -- an example of behavioral-based targeting -- which allows colleges to target Naviance users while they are on the platform.

The third dynamic in the EM industry has been the growth of test-optional admissions movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company [@blumenstyk_2019], failing [@molnar_2020a], and being acquired by Nexus Capital in 2024 [@knox_2024]. Over the last two decades, the testing companies developed new search filters that explicitly incorporated predictive analytics -- for example College Board's Geodemographic Segment filters [@RN4565] -- developed free college search engines, and entered the enrollment management consulting market more aggessively [@list_biz]. College Board has weathered the test-optional movement thanks to robust revenues from the Advanced Placement product, but the decline in PSAT/SAT/PreACT/ACT test-takers has undermined their oligopoly in the market for student list data.

The fourth -- and most important -- dynamic is the transformation of EM from an owner-operated to private-equity owned industry and the resulting eroding distinction between consultancy and software vendor. The 1980s and 1990s was a period of market entry in EM, when college admissions professionals or professors hung up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. Drawing from resource dependence theory, @list_biz describe the proliferation of horizontal acquisitions to grow market share and vertical acquisitions designed to make customers more dependent on particular recruiting products. After regulatory scrutiny pushed private-equity interests out of the for-profit college market [@RN5022], private equity investors found value in acquiring firms that provided services to direct providers. Direct providers are subsidized by state and federal funds (Title IV financial aid) and these funds are then paid to third-party vendors. Therefore, private equity ownership of direct providers and third-party vendors are both examples of government-subsidized rent-seeking.

EAB exemplifies the trend towards private equity ownership and industry concentration. EAB is the preeminent EM consultancy, offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB trace to Bill Royall, who founded Royall & Company in 1983 to provide direct marketing and fundraising for Republican political campaigns [@jump_2020]. By 1995 EM became their primary focus. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

As an owner-operated consultancy, Royall was the market leader in purchasing names from student list vendors on behalf of clients. Under Vista, EAB entered the market for student list data by completing a series of acquisitions that targeted the "leads" and "inquiries" stages of the enrollment funnel. In 2019, EAB acquired YouVisit, a leading provider of virtual tours. YouVisit is an engine for "inquiries" in that taking a virtual tour shows interest in a college and leaves a trail of contact information for subsequent recruiting interventions. In 2020, EAB acquired Cappex, a leading free college search engine, which is an engine for leads in that students provide contact details and other information. In 2021, EAB and PowerSchool -- the leading provider of K12 student information systems and a subsidiary of Vista -- jointly acquired the EM consultancy/software vendor Hobsons. PowerSchool acquired Hobson's Naviance product and its sister recruiting product Intersect and EAB paid PowerSchool to become the exlusive reseller of Intersect. <!--Finally, in 2022, EAB acquired Concourse, the leading "direct admissions" platform.--> EAB integrated these acquisitions to create the [Enroll360](https://eab.com/solutions/enroll360/) recruiting platform.^[The Enroll360 press release states, "We spent the last couple of years creating a connected recruitment ecosystem that allows enrollment leaders to keep pace with students pursuing increasingly digital journeys to college. This work led us to join forces with several leading companies: Cappex, Intersect, Wisr, and YouVisit\ldots Individually, each solution can solve important challenges at various stages of the enrollment funnel\ldots By bringing these capabilities together, our vision is to reinvent how enrollment leaders reach their goals [@koppenheffer_2021]."]

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50 per name). The goal was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board's *Student Search Service,* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and wheather students used Naviance to "research competitor institutions" [@markup_naviance]. A University of Utah procurment document states the Intersect subscription is necessary because "there is unique group of prospective students who are only in the PowerSchool Naviance platform" [@utah_procurement].

<!--# The marketing services being purchased operates within the PowerSchool Naviance platform, which is a proprietary system, that no other company has access to. There is unique group of prospective students who are only in the PowerSchool Naviance platform. We have spoken with several vendors where we can reach out to prospective students. Each vendor has their own proprietary website or database where select students can be reached.-->

An under-studied and under-regulated third-party EM industry has negative consequences for colleges and for students. Accross multiple student list products, we see a growing number of filters that recommend which prospects to target based on the beheavior of previous cohorts of students. The ACT/Encoura "Enrollment Predictor" filter enables colleges to filter prospects based on the predicted probability of enrollment, which is created based on analyzing the enrollment decisions of previous cohorts of students [@schmidt_2019]. College Board Geodemographic Segment filters categorizes each Census Tract into one of about two dozen clusters based on the college-going behavior of previous cohorts [@RN4565] and is based on the Claritas/PRIZM market segmentation system that categorized zip codes into groups useful for merchants [@RN4957]. An investor-facing @vista_eab_ml video promoting EAB's "Pipeline Analytics" product states, "we are using artificial intelligence and machine learning to help schools build list of students that may be more more likely to enroll and persist in their university." The product "identifies students that would be a good match for a school based on their similarity to students who have succeeded there before and then provides them \[the students\] information about those institutions so that they choose to apply and ultimately enroll." Each of these products are based on the underlying logic of correlations and homophily discussed by @RN4975 and applied by @RN4982. These products identify correlations in a historic snapshot of students and apply those relationships to make recommendations about which prospects to target in the present. As such, these technologies amplify the effect of previous inequality in student demand, which is itself a consequence of structural inequality in educational opportunity.

Market dynamics in the EM industry raise policy concerns about tuition prices, student choice, and competition. To the extent that colleges must buy expensive software to recruit prospects contained in an associated database, this expenditure will ultimately be passed on to students in the form of higher prices or reduced expenditure on education. Unfettered horizontal and vertical acquisitions in the EM industry causes the EM industry to become more concentrated. Smaller consultancies that do not control vertical inputs may be unable to compete with companies that have acquired proprietary pools of prospects. Fewer EM consultancies means less competition and higher consulting fees, which will be passed on to consumers. Finally, the blurring of the lines between consultancy and student list vendor suggests that large consultancies will funnel prospects only to those colleges that pay for subscriptions, raising policy concerns about student choice.
